Addressing selection bias and measurement error in COVID-19 case count data using auxiliary information

### 2022-03-21 write-up/letters/response-ref-1-v1.tex

\item {\it The AE also mentions a few comments as well:}
\begin{enumerate}
	\item {\it Title: As noted by one of the reviewers, the “paradox” is never discussed/defined in the manuscript. Consider changing to “Statistical estimation of Coronavirus Case Counts: Measurement…” or some alternative.}
	\vspace{5mm}

	We agree with the reviewer. This was not updated in the prior submission simply to ensure continuity of presentation (i.e., didn’t want to change from first submission). Here we replace with the title “Addressing selection bias and measurement error in COVID-19 case count data using auxiliary information”.

	\item {\it Organization: The figures and tables are not located anywhere near much of their introduction/discussion. For example, Table 1 was discussed on page 7 and is itself located on page 22. Consider the reorganization such that Figures and Tables are located near their discussion.}
	\vspace{5mm}

	We apologize.  We kept figures/tables related to the case study in Section 5 (the case study section) but can now see this led to confusion as we discuss them early in the paper. Tables and figures have been moved to be located much closer to their introduction.  \textcolor{red}{Need to do!}

	\item {\it Discussion on the Indiana Random Sample: On page 22, table 1, it appears that the random sample is not similar to the census estimates, potentially indicating non-response bias.}

    See our answer to Question 5a from Reviewer \#1. We now remark on the response rate in Section 5.  Specifically, we cite how Yiannoutsos et al. (2021) handle non-response in prevalence estimation.  We acknowledge the potential for a difference in characteristics between intended and final sample.

@@

	\item {\it Additionally, there may be significant potentially non-ignorable variables (e.g., political opinions and rural status) that drive non-response and are not presented. Is there any comment/discussion on the success of the randomization to be used as the “gold-standard” in this example? Much of the results appear to rely on the SRS and it ignores the potential for nonresponse bias.}

	We do not see random sampling as a ``gold-standard'' nor a panacea that will fix every issue.  Instead, we see it as a means to remove one very important source of potential bias in the estimation of population quantities of interest.  A key takeaway of our paper is that with access to \emph{only} a non-probabilistic sample, we know there is potential bias but have little hope of reducing said bias.  With a probability sample, we have means to reduce said bias.  Statisticians, epidemioligists, and government officials, and health policy experts can then focus on practical issues such as non-ignorable missingness.  If we thought, for example, political opinions and rural status are important predictors of missingness, then these should be collected as part of the random sampling protocol.  Note, even collecting strata information that is correlates with testing propensity (e.g., symptom status) in the non-probabilistic sample is insufficient to address selection bias.  This is not true for probability samples.  We have added this important discussion to Section 6.
\end{enumerate}

\end{enumerate}
\newpage
{\bf Response to Reviewer 1}
\vspace{5mm}

@@

	See answers to Questions 5a and 14a above.  The prevalence estimate comes from Yiannoutsos et al. (2021) who detail their adjustments for response rates in estimation of active prevalence.
	\vspace{5mm}
\end{enumerate}
\item {\bf Section 5.1.2 Time varying estimate}
\begin{enumerate}
	\item {\it 5.1.2. In the second approach, using hospitalization data to attempt to back-predict fever status seems highly likely to be subject to survivor bias as well as biases driven by differential access to care. Some discussion of the issues with this approach is warranted.}
	\vspace{5mm}

	We thank the reviewer for raising this issue.  We have added a remark in sction 5.1.2 that raises this as a limitation.  We restart here that the need for the approaches is due to the lack of availability of symptom status (which we know the state collects but has told us is unavailable due to data privacy / confidentiality issues).  Therefore, as a framework for health policy experts and government officials who have access to the complete government datasets (i.e., not just the public data), these steps will not be necessary.
	\vspace{5mm}
	\item {\it Also, focusing on the non-Hispanic, white male groups seems potentially to limit understanding of the scope of error. Although Indiana has a very high percent of white individuals, the individuals who present for testing (as well as those who test positive, are exposed in the workplace and home, and hospitalized or die) are disproportionately non-white. It is important to show the performance of the method in both white and non-white individuals because non-white individuals are the ones who are experiencing higher rates of illness and death and therefore will likely be more negatively impacted by biased estimates of disease prevalence.}
	\vspace{5mm}

	We acknowledge this shortcoming.  We now include versions of Figure 6 for other groups in Section H.2.1 of the supplementary materials with a discussion of the relative propensities.  Specifically, as the reviewer states above, individuals who identify as Hispanic, Male, and select `Some other race' were twice as likely to present for testing as non-Hispanic, white men -- empirically confirming that our data captures these systematic differences.
	\vspace{5mm}
\end{enumerate}
\item {\bf 5.2 SEIR model}
\begin{enumerate}
	\item {\it As stated above, some description of the assumptions related to contact patterns and transmission between and among age strata is needed. In addition, please clarify which compartments of the model the infection-fatality rate applies to (E \& I or I only), as well as how the model handles death from non-COVID causes from these and other model compartments.}
	\vspace{5mm}

@@

\vspace{5mm}
See our answer to Question 17b of Reviewer \#1. We acknowledge that these IFR estimates are likely biased and time-varying due to introduction of vaccines in 2020.  A sensitivity analysis focused on the issue of choosing an appropriate IFR and allowing for the IFR to be time-varying is now performed.
\textcolor{red}{To Do!}

\vspace{5mm}
\item {\it I found Figure 3 in the supplement [confusing]. Figure 3A shows the profile for a 35-44 white male with symptoms and COVID-19 contact and Figure 3B shows the profile for a 65-74 white female without symptoms nor COVID-19 contact – I would have expected the covariates between these (age, ethnicity, race, gender) to be the same rather than differ with the only thing differing being symptoms 1/ COVID-19 contact? As a minor note, the y-axis labels do not match for this plot.}

\vspace{5mm}
We have fixed Figure 3b to match on covariates and have matching y-axes.  See our answer to Question 16b from Reviewer \#1 on additional figures we are including in the revision.
\textcolor{red}{Need to do.}
\vspace{5mm}
\item {\it The model based estimates in Figure 8 seem so different from the other estimates that it is potentially implausible and suggests misspecification of the IFR. Is this potentially due to the age distribution of cases and the IFR collapsing choices? Perhaps as a sensitivity analysis different IFR values should be examined}

\vspace{5mm}
See our answer to Question 17b of Reviewer \#1 and Question 7 above. A sensitivity analysis focused on the issue of choosing an appropriate IFR is now performed.
\textcolor{red}{To Do!}
\vspace{5mm}
\item {\bf Minor comments:}
	\begin{enumerate}
		\item {\it Figure 3 mention “$\beta = 1.2$ (black)” – I’m not sure what “black” refers to?}
		\vspace{5mm}

### 2022-03-21 write-up/covid-supplement-writeup.tex
only seir fixes to go

\subsubsection{Propensity plots}

Test

\begin{figure}[!th]
\centering
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig1_supp4.png}
 \caption{Symptom Likelihood Given Negative Test}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
 \centering
\includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig2_supp4.png}
 \caption{Symptom Likelihood Given Positive Test}
\end{subfigure}
\caption{Likelihood of COVID-19 contact for Non-Hispanic, White Females}
\end{figure}

\begin{figure}[!th]
\centering
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig1_supp1.png}
 \caption{Symptom Likelihood Given Negative Test}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
 \centering
\includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig2_supp1.png}
 \caption{Symptom Likelihood Given Positive Test}
\end{subfigure}
\caption{Likelihood of COVID-19 contact for Non-Hispanic, White Females}
\end{figure}

\begin{figure}[!th]
\centering
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig1_supp2.png}
 \caption{Symptom Likelihood Given Negative Test}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
 \centering
\includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig2_supp2.png}
 \caption{Symptom Likelihood Given Positive Test}
\end{subfigure}
\caption{Likelihood of COVID-19 contact for Non-Hispanic, White Females}
\end{figure}

\begin{figure}[!th]
\centering
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig1_supp3.png}
 \caption{Symptom Likelihood Given Negative Test}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
 \centering
\includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig2_supp3.png}
 \caption{Symptom Likelihood Given Positive Test}
\end{subfigure}
\caption{Likelihood of COVID-19 contact for Non-Hispanic, White Females}
\end{figure}

>>>>

\subsubsection{Propensity plots}
\label{section:propplots}

Here we present the testing propensity for four strata: (1) non-Hispanic, white female, (2) non-Hispanic, African American male, (3) Hispanic, White male.   Strata (1) presented in Figure~\ref{fig:nonh-white-female} demonstrates a small gender difference in testing propensities.  Strata (2) and (3) presented in Figures~\ref{fig:nonh-aa-male} and~\ref{fig:h-white-male} demonstrates a lower rate of testing among African Americans and Hispanic white men compared to non-Hispanic, white men respectively (approximately 2 times lower). Strata (4) presented in Figure~\ref{fig:h-other-male} demonstrates a much higher testing rate for Hispanic men who select ``Some Other Race''.

\begin{figure}[!th]
\centering
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig1_supp4.png}
 \caption{Symptom Likelihood Given Negative Test}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
 \centering
\includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig2_supp4.png}
 \caption{Symptom Likelihood Given Positive Test}
\end{subfigure}
\caption{Likelihood of COVID-19 contact for Non-Hispanic, White Females}
\label{fig:nonh-white-female}
\end{figure}

\begin{figure}[!th]
\centering
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig1_supp1.png}
 \caption{Symptom Likelihood Given Negative Test}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
 \centering
\includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig2_supp1.png}
 \caption{Symptom Likelihood Given Positive Test}
\end{subfigure}
\caption{Likelihood of COVID-19 contact for Non-Hispanic, African American Males}
\label{fig:nonh-aa-male}
\end{figure}

\begin{figure}[!th]
\centering
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig1_supp2.png}
 \caption{Symptom Likelihood Given Negative Test}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
 \centering
\includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig2_supp2.png}
 \caption{Symptom Likelihood Given Positive Test}
\end{subfigure}
\caption{Likelihood of COVID-19 contact for Hispanic, White Males}
\label{fig:h-white-male}
\end{figure}

\begin{figure}[!th]
\centering
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig1_supp3.png}
 \caption{Symptom Likelihood Given Negative Test}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
 \centering
\includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig2_supp3.png}
 \caption{Symptom Likelihood Given Positive Test}
\end{subfigure}
\caption{Likelihood of COVID-19 contact for Hispanic, Males who select `Some Other Race'.}
\label{fig:h-other-male}
\end{figure}

### 2022-03-21 write-up/covid-writeup.tex
only seir fixes to go

\begin{remark}[An evolving pandemic]
This manuscript focuses on COVID-19 case count, testing, and death data collected from April through February 2021.  Numbers presented on disease dynamics are therefore based on the original strain.  Selection bias and measurement error persist in data arising from the delta and omicron strains and will likely persist for future variants. While not discussed in this manuscript, the framework presented will remain an appropriate tool for addressing selection bias and measurement error in these settings.
\end{remark}

@@

Here we provide the necessary background to understand COVID-19 diagnostic testing, its scientific use in managing the pandemic, and the data streams considered in this manuscript.

@@

 The primary goal of this manuscript is to produce accurate estimates of the population-level active infection rates over time using \emph{publicly available} viral testing data.  Secondary goals include estimation of rates of change and the effective reproduction number which characterize disease trajectory.
 These quantities are fundamental to public health policy and provide critical information on the presence and transmission of SARS-CoV-2.

@@

\begin{remark}[Public versus government datasets]
In September 2020, the author submitted a data request to the Indiana Data Hub (IDH).  Over the next three months, we worked with Indiana's COVID-19 response team to refine the request.  In December 2020, the request was officially approved and data streams were made publicly available on the IDH.  In this manuscript, we thus focus on this \emph{publicly-available data}, i.e., data  that anyone can download directly from the IDH. While the scientific community has contributed through independent COVID-19 observational and clinical studies, a significant component of public policy guidance has relied on testing and case count data, e.g., CDC and state guidelines based on test positivity rates and relative changes in the case counts over time.   Addressing selection bias and measurement-error in these public datasets is imperative in better informing public policy debate.

@@

 Comparing the mean-squared error (MSE) under a selection mechanism $\I$ with imperfect testing and SRS with perfect testing, we see that
 $$
 \frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]
 $$
 implying the error relative to SRS increases as a function of population size~\citep{Meng2018}. That two countries with the same testing strategy can yield wildly different estimates due to population size is too often ignored in current communications of case-count data.
 % Thus, conclusions drawn from observed case-counts may not just be wrong, but very wrong.

 A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' In Appendix~\ref{app:effss}, we show the effective sample size $n_{eff}$ can be bounded by
 $$
 \frac{f}{1-f} \times \frac{1}{E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
 $$
 Between April 25th to 29th 2020, Indiana performed $95,879$ tests.  Indiana's population is roughly $6.732$ million, so $f = 0.003$.  The active infection rate was estimated to be~$1.81\%$~\citep{Yiannoutsos2021} in this time interval. Recent studies have suggested 87\% sensitivity~\citep{Arevalo2020} and 97.6\% specificity~\citep{Cohen2020} are reasonable measurement error rates for RT-PCR tests. Supposing COVID-19 positive individuals are $1.5$ times more likely to get tested, then the effective sample size is $168$. Recent proposals~\citep{Siddarth2020} have argued for increased testing capacity, which may likely reduce the relative sampling rate.  Even if the relative sampling rate drops to $1.2$ and $f$ increases to $0.01$ then the effective sample size will increase to $1025$.  Thus the effective sample size even in optimistic scenarios is equivalent to a moderate random sample from the population.  Moreover, increased testing capacity may alter false positive and negative rates due to changes in sample collection quality, e.g., a testing center switches from nasal swab to oropharyngeal swabs or saliva specimen to speed up data collection.  Consider the case where $f$ increasing from $0.003$ to $0.01$ and is associated with the false negative rate rising from $13\%$ to $20$\%.  Then the effective sample size is $863$, representing a $5.1$ factor increase rather than the expected $6.1$ factor increase. See Section~\ref{section:effss} in the Supplementary Materials for additional effective sample size calculations.
 % \begin{wrapfigure}{r}{0.5\textwidth}

@@

 Figure~\ref{fig:testinglik_mainpaper} presents the testing propensity for non-Hispanic, white males across age strata and fever status using the second approach.  Note that the ratio of testing propensity within strata across fever status is time-varying, starting 50 times more likely in early April and dropping to ten times more likely by end of 2020.

@@

@@

 \textcolor{red}{I want to say that if i have many more random samples, then we should bias towards the random samples and not only smooth the means. But this is future work.}

>>>>

\begin{remark}[An evolving pandemic]
This paper focuses on COVID-19 case count, testing, and death data collected from April through February 2021.  Numbers presented on disease dynamics are therefore based on the original strain.  Selection bias and measurement error persist in data arising from the delta and omicron strains and will likely persist for future variants. While not discussed in this paper, the framework presented will remain an appropriate tool for addressing selection bias and measurement error in these settings.
\end{remark}

@@

Here we provide the necessary background to understand COVID-19 diagnostic testing, its scientific use in managing the pandemic, and the data streams considered in this paper.

@@

 The primary goal of this paper is to produce accurate estimates of the population-level active infection rates over time using \emph{publicly available} viral testing data.  Secondary goals include estimation of rates of change and the effective reproduction number which characterize disease trajectory.
 These quantities are fundamental to public health policy and provide critical information on the presence and transmission of SARS-CoV-2.

@@

\begin{remark}[Public versus government datasets]
In September 2020, the author submitted a data request to the Indiana Data Hub (IDH).  Over the next three months, we worked with Indiana's COVID-19 response team to refine the request.  In December 2020, the request was officially approved and data streams were made publicly available on the IDH.  In this paper, we thus focus on this \emph{publicly-available data}, i.e., data  that anyone can download directly from the IDH. While the scientific community has contributed through independent COVID-19 observational and clinical studies, a significant component of public policy guidance has relied on testing and case count data, e.g., CDC and state guidelines based on test positivity rates and relative changes in the case counts over time.   Addressing selection bias and measurement-error in these public datasets is imperative in better informing public policy debate.

@@

 Comparing the mean-squared error (MSE) under a selection mechanism $\I$ with imperfect testing and SRS with perfect testing, we see that
 $$
 \frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right].
 $$
 % implying the error relative to SRS increases as a function of population size~\citep{Meng2018}. That two countries with the same testing strategy can yield wildly different estimates due to population size is too often ignored in current communications of case-count data.
 % % Thus, conclusions drawn from observed case-counts may not just be wrong, but very wrong.
 A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' In Appendix~\ref{app:effss}, we show the effective sample size $n_{eff}$ can be bounded by
 $$
 \frac{f}{1-f} \times \frac{1}{E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
 $$
 Between April 25th to 29th 2020, Indiana performed $95,879$ tests.  Indiana's population is roughly $6.732$ million, so $f = 0.003$.  The active infection rate was estimated to be~$1.81\%$~\citep{Yiannoutsos2021} in this time interval. Recent studies have suggested 87\% sensitivity~\citep{Arevalo2020} and 97.6\% specificity~\citep{Cohen2020} are reasonable measurement error rates for RT-PCR tests. Supposing COVID-19 positive individuals are $1.5$ times more likely to get tested, then the effective sample size is $168$. Recent proposals~\citep{Siddarth2020} have argued for increased testing capacity, which may likely reduce the relative sampling rate.  Even if the relative sampling rate drops to $1.2$ and $f$ increases to $0.01$ then the effective sample size will increase to $1025$.  Thus the effective sample size even in optimistic scenarios is equivalent to a moderate random sample from the population.  Moreover, increased testing capacity may alter false positive and negative rates due to changes in sample collection quality, e.g., a testing center switches from nasal swab to oropharyngeal swabs or saliva specimen to speed up data collection.  Consider the case where $f$ increasing from $0.003$ to $0.01$ and is associated with the false negative rate rising from $13\%$ to $20$\%.  Then the effective sample size is $863$, representing a $5.1$ factor increase rather than the expected $6.1$ factor increase. See Section~\ref{section:effss} in the Supplementary Materials for additional effective sample size calculations which show that while~\citep{Meng2018} argues the relative error increases as a function of population size, our calculations indicate this is not true for fixed relative frequency~$f_1/f_0$.
 % \begin{wrapfigure}{r}{0.5\textwidth}

@@

 Figure~\ref{fig:testinglik_mainpaper} presents the testing propensity for non-Hispanic, white males across age strata and fever status using the second approach.  Note that the ratio of testing propensity within strata across fever status is time-varying, starting 50 times more likely in early April and dropping to ten times more likely by end of 2020.  Similar plots for other strata and a discussion of their relative testing propensities can be found in Section~\ref{section:propplots} of the supplementary materials.

@@

\textcolor{red}{
 While this paper emphasizes the importance of random sampling, we do not naively view it as a panacea. With access to \emph{only} a non-probabilistic sample, we have no means to reduce self-selection bias.  Adding a probability sample gives us the necessary auxiliary information to do so.  This leaves the data analyst more time to focus on statistical issues related to the random sample such as how to handle non-response bias.  Covariates that the scientific team thinks are correlated with non-response (e.g., political opinion) should be collected as part of the random sampling protocol and corrected for in the data analysis. On the other hand, collecting covariate information that correlates with testing propensity (e.g., symptom status) in the non-probabilistic sample is insufficient to address selection bias.
}

@@

 % \textcolor{red}{I want to say that if i have many more random samples, then we should bias towards the random samples and not only smooth the means. But this is future work.}
